What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Grammarly's proofreading for end of Section V (side effects)

The \rdyntrace contains low-level hooks capturing the life-cycle of an
environment and variables defined in it. We register callbacks in the tracer
that records each environment creation and removal with updates to its
bindings. From the recorded data, we ignore side effects coming from \eval
sites from unit testing frameworks.\footnote{In the corpus we have \c{RUnit,
testthat, tinytest} and \c{unitizer} unit testing frameworks.} They run the
testing code via \eval, and thus the results would be biassed because of the
high number of code in these tests (\CranRunnableCodeTestsRnd lines).

From the \packageNbrunsRnd programs we capture \SEAllRnd side effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side-effects caused by the R virtual machine implementation and are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator is saved and restored from
and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
packages. From these function, \SEFunsNighty are responsible for 90\% of side
effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object},
which represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a
function to have a given target environment class, all of its \eval sites that
do side effects must have the same target environment class The majority of the
\eval sites (\SESitesInOneClass) do all side effects consistently in one
environment class. The same happens on the function level. Almost half of the
sites (over a third of the functions) do side effects in either \emph{Local} or
\emph{Object} environments. This gives a ray of hope for the compiler. Even
though it is possible to do anything anywhere, the data suggest that most
side-effects are sane.

@@

Table~\ref{tab:se-types} shows the proportion of the different side effects
that were recorded. In terms of calls, we see primarily assignments and in
terms of site definitions. This is expected. A subsequent \eval call will turn
a definition into an update. The most dangerous side effect is variable removal
as it means that after a call to \eval some binding in some environments will
disappear. While rare, this happens, but the vast majority comes from a single
site called by the already mentioned \c{withr::execute\_handlers} function.
This makes sense as it is used to defer evaluation of an expression to after
the function exit and thus used for clean up. In the corpus, it is used almost
exclusively by the \c{tidyselect} package for removing the reference to the
current quosure environment while interpreting a data frame column
selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}

Looking closely at the value types in variable updates, we observe that the
majority of \eval sites involve basic R vectors (\SEBasicTypeRatio) and lists
(\SEListTypeRatio). From the compiler perspective, we would like to know how
many sites changes function bindings. In the corpus, this happens in only
\SEClosureType sites, \SEClosureTypeLocal of them do that in the local
environment. We have manually inspected a few of these sites, but except for
manual injection of parameters into a \c{model.frame} execution environment, we
did not find a common use case.

@@

\mypara{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope.
In R, given the support for first-class environments, it can happen anywhere
making it \eval more dangerous than it already is. The data however suggest
that first, side-effects from \eval is not as wide spread as in the case of
JavaScript and that over half of them happen in a predictable environment.

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are the tools given to
developers to this end. This API is slightly more complex than just
passing a string, it is conceivable that this may discourage some
casual users. \Eval is also being used to reduce boilerplate code and
provide convenience features for programmers. We give examples for the minimized expressions of Section~\ref{sec:minimized} and then dig into high-level \eval design patterns.

>>>>

The \rdyntrace contains low-level hooks capturing the life-cycle of an
environment and variables defined in it. We register callbacks in the tracer
that records each environment creation and removal with updates to its
bindings. From the recorded data, we ignore side effects coming from \eval
sites from unit testing frameworks.\footnote{In the corpus, we have \c{RUnit,
testthat, tinytest} and \c{unitizer} unit testing frameworks.} They run the
testing code via \eval, and thus the results would be biased because of the
high number of code in these tests (\CranRunnableCodeTestsRnd lines).

From the \packageNbrunsRnd programs we capture \SEAllRnd side effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side effects caused by the R virtual machine implementation and are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator, is saved and restored from
and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
packages. From these functions, \SEFunsNighty are responsible for 90\% of side
effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally, \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object},
which represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a
function to have a given target environment class, all of its \eval sites that
do side effects must have the same target environment class The majority of the
\eval sites (\SESitesInOneClass) do all side effects consistently in one
environment class. The same happens on the function level. Almost half of the
sites (over a third of the functions) do side effects in either \emph{Local} or
\emph{Object} environments. This gives a ray of hope for the compiler. Even
though it is possible to do anything anywhere, the data suggest that most
side-effects are sane.

@@

Table~\ref{tab:se-types} shows the proportion of the different side effects
that were recorded. In terms of calls, we see assignments primarily and in
terms of site definitions. This is expected. A subsequent \eval call will turn
a definition into an update. The most dangerous side effect is variable removal,
as it means that after a call to \eval some binding in some environments will
disappear. While rare, this happens, but the vast majority comes from a single
site called by the already mentioned \c{withr::execute\_handlers} function.
This makes sense as it is used to defer evaluation of an expression to after
the function exit and thus used for clean up. In the corpus, it is used almost
exclusively by the \c{tidyselect} package for removing the reference to the
current quosure environment while interpreting a data frame column
selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}

Looking closely at the value types in variable updates, we observe that the
majority of \eval sites involve basic R vectors (\SEBasicTypeRatio) and lists
(\SEListTypeRatio). From the compiler perspective, we would like to know how
many sites changes function bindings. In the corpus, this happens in only
\SEClosureType sites; \SEClosureTypeLocal of them do that in the local
environment. We have manually inspected a few of these sites, but except for
manual injection of parameters into a \c{model.frame} execution environment, we
did not find a common use case.

@@

\mypara{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope.
In R, given the support for first-class environments, it can happen anywhere, 
making it \eval more dangerous than it already is. However, the data  suggest
that first, side-effects from \eval are not as widespread as in the case of
JavaScript and that over half of them happen in a predictable environment.

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute}, and \c{eval} are the tools given to
developers to this end. This API is slightly more complex than just
passing a string, it is conceivable that this may discourage some
casual users. \Eval is also being used to reduce boilerplate code and
provide convenience features for programmers. We give examples for the minimized expressions of Section~\ref{sec:minimized} and then dig into high-level \eval design patterns.
